Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2021-12-23 paper/paper.tex
update

\blue{independence polynomial at -1~\cite{Bousquet2008,Levit2009},
two motivations of computing independence polynomial: 1. Lee-Yang zero~\cite{Lee1952,Yang1952}, 2. Lov{\'a}sz local lemma~\cite{Scott2005}.
}
\section{Introduction}
In graph theory and combinatorial optimization, there are many interesting hard computational problems concerning properties of independent sets.
%An independent set of a graph is a subset of its vertices that no two vertices in this set are adjacent to each other.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$. %there is no edge connecting $u$ and $v$ in $G$.
The primary insterest of many computational complexity theorists is finding the maximum size of such vertex sets $\alpha(G) \equiv \max_{I}|I|$ and one of the sets of this size.
Finding $\alpha(G)$ is a hard computational problem; moreover, it is NP-hard to even approximate $\alpha(G)$ within a factor $|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$ ~\cite{Hastad1996}.
Naive exhaustive search for an MIS requires computing time $O\left(2^{|V|} \right)$.
People developed more efficient exact algorithms for finding MISs, including the branching algorithms~\cite{Tarjan1977,Robson1986} and dynamic programming.
The branching algorithms can reduce the base of the exponential time scaling to, e.g., $1.1996^{|V|} {|V|}^{O(1)}$~\cite{Xiao2017}, 
while dynamic programming approaches~\cite{Courcelle1990, Fomin2013} works better for graphs with a small treewidth $\text{tw}(G)$;
these methods can produce algorithms of complexity $O(2^{\text{tw}(G)} \text{tw}(G) |V|)$.
Although finding MIS is hard, people keep chanllenging themselves to find better algorithms for solving the MIS problem;
not only for it has a wide range of applications in scheduling, logistics, social network analysis, bioinformatics, wireless networks and telecommunication,
map labelling, computer vision, etc.~\cite{Butenko2003, Wu2015}, but also for it is a well know NP-complete problem that can be mapped from
 to many other important combinatorial optimization problems in polynomial time such as circuit satisfiability~\cite{}, the maximum clique problem and the maximum vertex cover~\cite{Moore2011} problems.

@@

\section{Tensor networks}
Tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts,
it can be viewed as a generalization of matrix multiplication to multiple tensors contraction.
People often use Einstein's notation to represent a tensor network, e.g. the matrix multiplication between two matrices $A$ and $B$ can be represented in Einstein's notation as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in subscripts to represent a degree of freedom.
We enumerate these degree of freedoms and accumulate the product of tensor elements to the output tensor.
In the standard notation of Einstein's summation or tensor network in physics, each index appears precisely twice.
Hence people can represent a tensor network as a simple graph,
where a tensor is a vertex and two tensors sharing the same label are connected.
In the main text, we do not have this restriction. An index can appear arbitrary times.
The graphical representation of our generlized tensor network is hypergraph, in which an edge (label) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors - a high dimensional equivalence of identity matrix.
However, introducing $\delta$ tensors can potentially increase the contraction complexity a lot. We illustrate this point in \App{app:tensorbad}.
\blue{introduce contraction orders}

@@

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
            \textbf{element type}     & \textbf{purpose} \\
            {$\mathbbm{R}$}     & {real number for counting all independent sets} \\
            {$\mathbbm{C}$}     & {complex number for fitting the independence polynomial with fast Fourier transform} \\
            {$\text{GF}(p)$} (\Eq{eq:finitefield}) & {number on finite field algebra for fitting the independence polynomial exactly using number theory} \\
            {T} (\Eq{eq:tropical})    & {tropical number for finding the independence number} \\
            {P1} (\Eq{eq:countingtropical})     & {polynomial truncated to largest order, i.e. the tropical number with counting field for finding the independence number and the number of MISs} \\
            {P2} (\Eq{eq:max2poly})     & {counting MISs and independent sets of size $\alpha(G)-1$} \\
            {PN} (\Eq{eq:polynomial})     & {computing the independence polynomial exactly} \\
            {P1+S1}     & {replacing the counting field in P1 to configuration sampler defined in $\Eq{eq:singleconfig}$ for finding the independence number and one MIS} \\
            {P1+SN} (\Eq{eq:countingtropicalset})    & {replacing the counting field in P1 to configuration sampler defined in $\Eq{eq:set}$ for finding the independence number and enumeration of all MISs} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and their purposes in calculating various independent set properties.}\label{tbl:generictypes}
\end{minipage}
\red{we should include maximal independence polynomial and counting, enumeration in this table as well.}
\blue{I do not think so. Network structure and element types are two degree of freedoms,
maximal IS is different network structure. We can use the above algebra in any network structure.}
\end{table}

>>>>

\blue{independence polynomial at -1~\cite{Bousquet2008,Levit2009},
two motivations of computing independence polynomial: 1. Lee-Yang zero~\cite{Lee1952,Yang1952}, 2. Lov{\'a}sz local lemma~\cite{Scott2005}.
}
\section{Introduction}
In graph theory and combinatorial optimization, there are many interesting hard computational problems concerning properties of independent sets.
%An independent set of a graph is a subset of its vertices that no two vertices in this set are adjacent to each other.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$. %there is no edge connecting $u$ and $v$ in $G$.
The primary insterest of many computational complexity theorists is finding the maximum size of such vertex sets $\alpha(G) \equiv \max_{I}|I|$ and one of the sets of this size.
Finding $\alpha(G)$ is a hard computational problem; moreover, it is NP-hard to even approximate $\alpha(G)$ within a factor $|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$ ~\cite{Hastad1996}.
Naive exhaustive search for an MIS requires computing time $O\left(2^{|V|} \right)$.
People developed more efficient exact algorithms for finding MISs, including the branching algorithms~\cite{Tarjan1977,Robson1986} and dynamic programming.
The branching algorithms can reduce the base of the exponential time scaling to, e.g., $1.1996^{|V|} {|V|}^{O(1)}$~\cite{Xiao2017}, 
while dynamic programming approaches~\cite{Courcelle1990, Fomin2013} works better for graphs with a small treewidth $\text{tw}(G)$;
these methods can produce algorithms of complexity $O(2^{\text{tw}(G)} \text{tw}(G) |V|)$.
People are eager to find better algorithms for solving the MIS problem;
not only for it has a wide range of applications in scheduling, logistics, social network analysis, bioinformatics, wireless networks and telecommunication,
map labelling, computer vision, etc.~\cite{Butenko2003, Wu2015}, but also for it is a well know NP-complete problem that can be mapped from
 to many other important combinatorial optimization problems in polynomial time such as circuit satisfiability~\cite{}, the maximum clique problem and the maximum vertex cover~\cite{Moore2011} problems.

@@

\section{Tensor networks}
Tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts,
it can be viewed as a generalization of matrix multiplication to multiple tensors contraction.
People often use Einstein's notation to represent a tensor network, e.g. the matrix multiplication between two matrices $A$ and $B$ can be represented in Einstein's notation as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in subscripts to represent a degree of freedom.
We enumerate these degree of freedoms and accumulate the product of tensor elements to the output tensor.
In the standard notation of Einstein's summation or tensor network in physics, each index appears precisely twice.
Hence people can represent a tensor network as a simple graph,
where a tensor is a vertex and two tensors sharing the same label are connected.
In the main text, we do not have this restriction. An index can appear arbitrary times.
The graphical representation of our generlized tensor network is hypergraph, in which an edge (label) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors - a high dimensional equivalence of identity matrix.
However, introducing $\delta$ tensors can potentially increase the contraction complexity a lot. We illustrate this point in \App{app:tensorbad}.
\blue{Is it better to introduce the contraction orders here or in example~\ref{eg:tensorcontraction}?}

@@

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
            \textbf{element type}     & \textbf{purpose} \\
            {$\mathbbm{R}$}     & {real number for counting independent sets} \\
            {$\mathbbm{C}$}     & {complex number for fitting the independence polynomial with fast Fourier transform} \\
            {$\text{GF}(p)$} (\Eq{eq:finitefield}) & {finite field algebra for fitting the independence polynomial exactly using number theory} \\
            {T} (\Eq{eq:tropical})    & {tropical number for finding the independence number} \\
            {P1} (\Eq{eq:countingtropical})     & {polynomial truncated to largest order, i.e. the tropical number with counting field for finding the independence number and the number of MISs} \\
            {P2} (\Eq{eq:max2poly})     & {polynomial truncated to largest two orders for counting MISs and independent sets of size $\alpha(G)-1$} \\
            {PN} (\Eq{eq:polynomial})     & {polynomial number for computing the independence polynomial directly} \\
            {P1+S1}     & {P1 with configuration sampler (\Eq{eq:singleconfig}) as its counting field for finding the independence number and one MIS} \\
            {P1+SN} (\Eq{eq:countingtropicalset})    & {P1 with configuration enumerator (\Eq{eq:set}) as its counting field for finding the independence number and enumeration of all MISs} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and their purposes in calculating various independent set properties.}\label{tbl:generictypes}
\end{minipage}
\red{we should include maximal independence polynomial and counting, enumeration in this table as well.}
\blue{I do not think so. Network structure and element types are two degree of freedoms,
maximal IS is different network structure. We can use the above algebra in any network structure.}
\end{table}
